fof(incidence1, axiom,((
	point(A)
	& point(B)
	& line(C)
	=> on_line(A,B,C)))).

fof(incidence2, axiom,((
	line(C)
	& line(D)
	& on_line(A,B,C)
	& on_line(A,B,D)
	=> equals(C,D)))).

fof(incidence4, axiom,((
	line(D)
	& plane(E)
	& on_line(A, B, D)
	& ~on_line(B, C, D)
	=> on_plane(A, B, C, E)))).

fof(incidence5, axiom,((
	line(D)
	& plane(E)
	& plane(F)
	& on_line(A, B, D)
	& ~on_line(B, C, D)
	& on_plane(A, B, C, E)
	& on_plane(A, B, C, F)
	=> equals(E,F)))).

fof(incidence6, axiom,((
	line(D)
	& plane(E)
	& on_line(A,B,D)
	& on_line(L,M,D)
	& on_plane(A,B,X,E)
	=> on_plane(L,M,X,E)))).

#D and F existential here
fof(incidence7, axiom,((
	plane(A)
	& plane(B)
	& on_plane(C, D, E, A)
	& on_plane(C, F, G, B)
	=> equals(D, F)))).

fof(order1, axiom,((
	point(A)
	& point(B)
	& point(C)
	& line(D)
	& between(B, A, C)
	=> between(B, C, A)
	& on_line(B, C, D)
	& on_line(A, C, D)))).

fof(order2, axiom,((
	point(A)
	& point(B)
	& point(C)
	& line(D)
	& on_line(A,C,D)
	=> between(B,A,C)
	& on_line(A,B,D)
	& on_line(B,C,D)))).

fof(congruence1, axiom,((
	point(A)
	& point(_A)
	& point(B)
	& point(_B)
	& line(C)
	& on_line(A,B,C)
	& on_line(A,_A,C)
	=> equals(dist(A,B),dist(_A,_B))))).

fof(congruence2, axiom,((
	dist(A,B)
	=> dist(B,A)))).

fof(congruence3, axiom,((
	point(A)
	& point(B)
	& point(C)
	& point(D)
	& point(E)
	& point(F)
	& equals(dist(A,B),dist(C,D))
	& equals(dist(C,D),dist(E,F))
	=> equals(dist(A,B),dist(C,D))))).

fof(congruence4, axiom,((
	point(H)
	& point(K)
	& angle(H,K)
	=> angle(K,H)))).

fof(congruenceFor3, axiom,((
	point(H)
	& point(K)
	& point(J)
	& angle(H,K,J)
	=> angle(J,K,H)))).

fof(congruence5, axiom,((
	point(H)
	& point(K)
	& point(J)
	& point(L)
	& point(Q)
	& point(R)
	& equals(angle(H,K),angle(J,L))
	& equals(angle(J,L),angle(Q,R))
	=> equals(angle(H,K),angle(Q,R))))).

fof(congruence6, axiom,((
	triangle(A,B,C)
	& triangle(_A,_B,_C)
	& equals(dist(A,B),dist(_A,_B))
	& equals(dist(A,C),dist(_A,_C))
	& equals(angle(B,A,C),angle(_B,_A,_C))
	=> equals(angle(A,B,C),angle(_A,_B,_C))))).
