%!TEX root = ../username.tex
\chapter{Previous Work}\label{text}                                                                   
\section{Natural Language Processors}                                                  
Natural language processing has been an area of interest in computer science for the past 
60 years. Initially, the subject was limited to machine translation, with the Georgetown e
xperiment in 1954 being an early foray into the field. In this experiment over sixty Russi
an sentences were translated to English entirely automatically. The results were promising
 enough that the researchers anticipated automatic translation being solved within the nex
t five years. \cite{Hutchins} Progress was, obviously, much slower than this.             
                                                                                          
Funding was cut when the program ran over a decade, and progress in the field slowed. Two 
successful systems were created in the 1960s, SHRDLU and ELIZA.                           
                                                                                          
\subsection{SHRDLU}                                                                    
SHRDLU was a querying system which created a small ``Blocks World'' which was populated by
 cones, spheres, cubes, and other geometric shapes of various sizes and colors. \cite{wino
grad} The user was able to instruct SHRDLU on how to move these objects around simply by s
pecifying the shape in addition to its color or size. SHRDLU also had a simple memory syst
em, allowing for reference to recently interacted with objects. This memory also allowed f
or SHRDLU to be queried on what she had previously done. SHRDLU's world contained basic ph
ysics, allowing for the program to describe what was possible in the world and what was no
t. The final major feature was to remember the name a user gave an object or collection of
 objects. From this, it was possible to more easily instruct SHRDLU.                      
                                                                                          
\subsection{ELIZA}                                                                     
                                                                                          
{Include a sample dialogue, and a flowchart of ELIZA's function}
