%!TEX root = ../username.tex
\chapter{Introduction}\label{intro}
ver the past decade, natural language processors have become more and more commonplace. I
n late 2011, Apple introduced their digital assistant, Siri, to iOS. Siri was the successo
r to voice control, an iOS feature that allowed the user to interact with their phone usin
g their voice, though in both an extremely limited fashion and rather unreliably. Siri was
 powered by a much better voice recognition model, and not only could process many command
s into action, but could even ask for clarification based on context. Since this time, Goo
gle introduced the Google Assistant to its Android and ChromeOS platforms, Amazon released
 smart home products powered by Alexa, and Siri has been ported to MacOS and WatchOS. All 
of these digital assistants rely on understanding the commands of users to be useful, and 
the only way this is possible is through natural language processing.                     
                                                                                          
From Q2 2017 to Q2 2018, smart speakers, the quintessential aspect of the smart home to ma
ny, had shipments grow 187\%. \cite{canalys} %https://www.canalys.com/newsroom/global-smar
t-speaker-shipments-grew-187-year-on-year-in-q2-2018-with-china-the-fastest-growing-market
                                                                                          
This growth is expected to continue in the coming years, with smart speakers and the smart
 home technology they may control becoming an enormous industry. As this industry expands,
 there will be greater and greater access to these systems among the general population. W
hat if it was possible for a student to ask their smart screen why a mathematical claim wa
s true, and that system could respond by displaying the proof? Smart devices offer more th
an simply convenience, they could become an amazing source of educational opportunities wh
ich are not otherwise available for many.                                                 
                                                                                          
While we will not be using smart devices in this project, we will be creating a front end 
interface for students to more easily investigate mathematical questions, without the barr
ier of language.                                                                          
                                                                                          
\section{Problem Statement}                                                            
Automated theorem proving systems are a powerful tool which can allow a student to check t
heir work or a researcher to prove a claim which could be extremely time consuming if done
 by hand. Unfortunately, these systems are not trivial to use, with the use of any such sy
stem requiring the user to learn the proper method for input, which will also differ betwe
en different such systems. As a result of this, it can be easy to be put off of these syst
ems, and once one learns one such system, it is easy to be stuck using what is already kno
wn, even if better options may be available. A wonderful solution to this problem would be
 leveraging computational power to convert a human language request, i.e. ``Prove that the
 length of the third side of a triangle is determined by the two other sides and the angle
 between them'' into a language the proving system will understand.

 \section{Project Aims}                                                                 
This project has two separate but interconnected goals; the production of a natural langua
ge processor, and the creation of a basic automated prover. The processor will:           
\begin{itemize}                                                                           
\item{Serve as a front end for any automated prover.}                                     
\item{Be fully functional without any understanding of symbolic notation or programming}  
\item{Be modular enough to become the front end of any automated proving system.}         
\item{Accurately translate standard English to symbolic notation.}                        
\end{itemize}                                                                             
                                                                                          
The prover will have far more modest aims:                                                
\begin{itemize}                                                                           
\item{Understand basic questions in a limited scope, such as geometry or arithmetic.}     
\item{Take symbolic input, and return the truth value of said statement.}                 
\item{Display the steps of the proof to the end user.}                                    
\end{itemize}                                                                             
                                                                                          
As the two pieces are separate, the processor will not be limited by the prover, and may i
nstead be joined with a more robust system by specifying what said system is. This will be
 necessary to convert the request properly, so that the processor will know which symbolic
 language to translate to.                                                                
\subsection{Natural Language Processor}                                                
The processor will take a user's input in English, and determine the goal of the user base
d on the content of that input. The user will not be required to know anything, other than
 that the system is intended for mathematical queries and will be able to do nothing else.
 Once the input has been analyzed, the processor will convert it to the proper symbolic la
nguage for the automated prover.                                                          
\subsection{Automated Theorem Prover}                                                  
                                                                                          
The prover will take a mathematical claim, in a symbolic notation, and apply axioms, theor
ems, and other knowledge as necessary to prove or disprove the claim. If the prover is abl
e to use theorems in its proofs, it will be barred from using them in such a way as to giv
e the appearance of circular logic. For example, when a user asks for a proof of the Pytha
gorean Theorem, the prover will be disallowed from using the Pythagorean Theorem as its ju
stification for the truth of the theorem. %We can do this by preventing the prover from us
ing single step proofs except in the case of an appeal to an axiom!                       
Should the user ask for a proof of an axiom, it will be permissible to simply point to the
 axiom. The prover should not and will not be expected to explain an axiom to a user. 

